A Triple Origin for Twin Blue Stragglers in Close Binaries

### 2019-01-14 BSS7782.tex
updated text and figures

We propose a formation channel for Binary 7782, and compact double BS
binaries in general, which involves mass transfer from an outer
tertiary companion on to an inner (MS-MS) binary.  In
section~\ref{sect:dyn}, we constrain the range of initial (i.e., pre-mass
transfer) orbital parameters for a hypothetical outer tertiary
companion, using a combination of dynamical and stellar
evolution-based constraints.  In Section~\ref{sims} we present the
numerical simulations used to study the mass transfer process in our
hypothetical triple system, computed using the
\texttt{AMUSE}\cite{AMUSE} software package, in order to study the
evolution of the inner and outer orbital parameters during mass
transfer.  We summarize and discuss the implications of our results
for compact double BS binaries and, more generally, mass transfer in
stellar triples in Section~\ref{sect:discussion}.

@@

Consider a hierarchical triple system with component masses $m_{\rm
  1}$ and $m_{\rm 2}$ for the inner binary, and mass $m_{\rm 3}$ for
the outer tertiary companion.  The inner and outer binary orbital
semi-major axes are denoted a$_{\rm in}$ and a$_{\rm out}$,
respectively.  For clarity we now assume both orbits, the inner as
well as the outer, to have a small eccentricity and low inclination.
These assumptions are also supported by the population of observed
low-mass triples \cite{2010yCat..73890925T,moe18}.  This initial
configuration for our assumed formation scenario for Binary 7782,
described below, is depicted in Figure~\ref{fig:fig1}.

@@

We consider a scenario where the outer star is filling its Roche lobe
and transfers mass to the inner binary.  The mass transfer stream
gathers at the circularization radius a$_{\rm c}$, and forms a circumbinary disk
around the inner binary.  Using conservation of angular momentum, we
equate the specific angular momentum of the accreted mass at the inner
Lagrangian point of the (outer) donor star to the final specific
angular momentum of the accretion stream at the circularization radius
about the inner binary, this results in
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, $a_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at $a_{\rm c}$.  The distance
from the center of mass corresponding to the outer tertiary companion
defined by the Roche lobe is given by Equation 2 in
\citep{1983ApJ...268..368E}.  Combining Equation 2 in \citet{1983ApJ...268..368E} (with
mass ratio q $=$ m$_{\rm 3}$/(m$_{\rm 1} +$m$_{\rm 2}$)) with
Equation~\ref{eqn:specangmom1}, we can solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

Figure~\ref{fig:fig2} shows the parameter space in the P$_{\rm
  out}$-P$_{\rm in}$-plane for Binary 7782.  We assume initial
component masses of $m_{\rm 1} = 1.1$ M$_{\rm \odot}$ and $m_{\rm 2}
= 0.9$ M$_{\rm \odot}$ for the inner binary components, and $m_{\rm 3}
= 1.4 $M$_{\rm \odot}$ for the outer tertiary.  We compare the
circularization radius to the semi-major axis of the inner binary, for
which we require $a_{\rm c} > a_{\rm in}$, after folding in all
constraints from the requirements for dynamical stability (listed in
the caption of Figure~\ref{fig:fig2}), an outer tertiary that is Roche
lobe-filling, and a dynamically hard outer tertiary orbit.  Note that
the range of plotted orbital periods P$_{\rm in}$ corresponding to a
contact state for the inner binary lies outside the range of plotted
values for $P_{\rm in}$ (for components with radii of roughly 1
R$_{\odot}$), since it does not contribute significantly to
constraining the outer orbital properties of a hypothesized outer WD
tertiary).  The thick horizontal solid red line shows the allowed range
of outer semi-major axes, after folding in all of the aforementioned
criteria.  As is clear, this makes a relatively narrow prediction for
the allowed ranges of outer tertiary orbits, namely 2.2 $\times$
10$^{2}$ days $\le$ P$_{\rm out}$ $\le$ 1.1 $\times$ 10$^3$ days, for
our assumed final donor mass.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_M14MSun_outerorbit.pdf}
  \caption{Giant radius as a function of the mass of the core of the
    Roche-lobe filling outer star (dark blue curve).  Here we adopted
    a donor mass of 1.4\,\MSun, but for an 1.2\,\MSun\, donor curve is
    quite similar (see dotted dark-blue curve).  The red squares in
    the curve show the parameters for which we performed more detailed
    gravitational-hydrodynamical simulations (see \S\,\ref{sims}).
    The horizontal dashed line shows the orbital separation of the
    observed twin blue stragger 7782.  The initial triple in which it
    possibly formed must at least have been dynamically stable. The
    minimal orbital separation for the inner binary for which the
    triple is stable is given by the lower green colored curve.
    Donors which are smaller than about 100\,\RSun\, (dark blue curve
    indicated with a$_{\rm stable}$) result in a dynamically unstable
    triple. The minimal core mass associated with a stable triple is
    then indicated by the left-most vertikcal dotted line.  The
    orbital separation at which the donor star overfills its Roche
    lobe is indicated with the light-blue curve. The top curve (dark
    red) shows an estimate of the final orbit separation of the outer
    star, and therefore of the final orbit of the white dwarf around
    the inner twin blue straggler.  For core masses $\apgt
    0.5$\,\MSun\, the final orbital separataion, after mass transfer,
    is smaller than the initial orbit.  Here we adopted an initial
    inner binary mass of (1.0+0.9)\,\MSun\, and a final twin-blue
    straggler mass of (1.4+1.4)\,\MSun.
\label{fig:tertiarymass_vs_size}}
\end{figure}

In Fig.\,\ref{fig:tertiarymass_vs_size} we present several
characteristic sizes and orbital separations for the problem at hand.
The red curves show the stellar radius upon reaching the first giant
branch (thick curve) and the asymptotic giant branch (thin curves).
The blue curves show the orbital separation at the moment of
Roche-lobe overflow for each of the aforementioned giants, assuming a
circular orbit and a total inner binary mass of $0.9 + 1.1$
M$_{\odot}$ $=$ 2.0 M$_{\odot}$.  The green curves show for these
assumptions the maximum orbital separation for a stable inner binary
pair.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_BBSS_gas_M14Msun_A010au_t0002.pdf}
\caption{Top view of the simulated triple system in which the outer
  star over-fills its Roche-lobe (represented by 80000\,sph particles
  and a core particle of 0.4\,\MSun (blue bullet). The two companion
  stars are represented as red bullets (to the right).
\label{fig:topview_at_t0}}
\end{figure}

@@

\begin{figure}[ht!]
  \includegraphics[width=0.49\columnwidth]{fig_BBSS_Lgas_M14Msun_A010au_t3360.pdf}
~  \includegraphics[width=0.51\columnwidth]{fig_BBSS_gas_M14Msun_A010au_t3360.pdf}
\caption{Same as Fig.\,\ref{fig:topview_at_t0} but at an age of $t
  \simeq 996days$. Left panel shows the equipotential surface of the
  triple overplotted with the gas distribution, the right panel shows
  just the gas and the stars as bullets.
\label{fig:topview_at_t1000day}}
\end{figure}

>>>>

We propose a formation channel for Binary 7782, and compact double BS
binaries in general, which involves mass transfer from an outer
tertiary companion on to an inner binary composed to two main-sequence
stars.  In section~\ref{sect:dyn}, we constrain the range of initial
(i.e., pre-mass transfer) orbital parameters for a hypothetical outer
tertiary companion, using a combination of dynamical and stellar
evolution-based constraints.  In Section~\ref{sims} we present the
numerical simulations used to study the mass transfer process triple
system with orbital parameters that, according to our expectations,
for promising progenitors for the twin BS 7782.  The calculations are
performed using the Astrophysical Multipurpose Software Environment
(\texttt{AMUSE} for short) \cite{AMUSE} with a combination of stellar
evolution, hydrodynamical and gravitational simulations.  with these
calculations we further constrain the possible range of initial
parameters that naturally lead to twin BS with orbital parameters
similar to the 7782 system.  We summarize and discuss the implications
of our results for compact double BS binaries and, more generally,
mass transfer in stellar triples in Section~\ref{sect:discussion}.

@@

In our scenario we start with a binary star with component masses
$m_{\rm 1}$ and $m_{\rm 2}$ that is orbited tertiary of mass
$m_3$. The inner and outer binary orbital semi-major axes are denoted
a$_{\rm in}$ and a$_{\rm out}$, respectively.  For clarity we now
assume both orbits, the inner as well as the outer, to have negligible
eccentricity and low inclination.  These assumptions are also
supported by the population of observed low-mass triples
\cite{2010yCat..73890925T,moe18}.  This initial configuration for our
assumed formation scenario for Binary 7782, described below, is
depicted in Figure~\ref{fig:fig1}.

@@

According to our scenario $m_3 > m_1 > m_2$ and the outer orbit is
sufficiently small that the tertiary star is filling its Roche lobe
and transfers mass to the inner binary before it leaves the asumptotic
giant branch. We constrain the inner orbit by requiring the triple
system to be dynamicall stable, for which we adopt eq.\,90 in
\cite{2001MNRAS.321..398M}.  While transferring mass, the accretion
stream gathers around the inner binary at the circularization radius
a$_{\rm c}$, and forms a circumbinary disk.  Using conservation of
angular momentum, we equate the specific angular momentum of the
accreted mass at the inner Lagrangian point of the (outer) donor star
to the final specific angular momentum of the accretion stream at the
circularization radius about the inner binary, this results in
\begin{equation}
\label{eqn:specangmom1}
v_{\rm orb,3}(a_{\rm out} - R_{\rm L}) = v_{\rm orb,c}a_{\rm c},
\end{equation}
where R$_{\rm L}$ is the radius of the Roche lobe of the outer
tertiary companion, $a_{\rm c}$ is the semi-major axis of the orbit
about the inner binary corresponding to the circularization radius and
v$_{\rm orb,c}$ is the orbital velocity at $a_{\rm c}$.  The distance
from the center of mass corresponding to the tertiary defined by the
Roche lobe is given by Equation 2 in \citep{1983ApJ...268..368E}.
Combining Equation 2 in \citet{1983ApJ...268..368E} (with mass ratio q
$=$ m$_{\rm 3}$/(m$_{\rm 1} +$m$_{\rm 2}$)) with
Equation~\ref{eqn:specangmom1}, we solve for the circularization
radius as a function of a$_{\rm out}$ and the assumed stellar masses:
\begin{equation}
\label{eqn:ac}
a_{\rm c} = a_{\rm out}(1 - R_{\rm L}).
\end{equation}
In order for a circumbinary disk to form around the inner binary, we
require that a$_{\rm in} <$ a$_{\rm c}$.

Figure~\ref{fig:fig2} shows the parameter space in the P$_{\rm
  out}$-P$_{\rm in}$-plane for Binary 7782.  We assume initial
component masses of $m_{\rm 1} = 1.1$ M$_{\rm \odot}$ and $m_{\rm 2} =
0.9$ M$_{\rm \odot}$ for the inner binary components, and $m_{\rm 3} =
1.4 $M$_{\rm \odot}$ for the outer tertiary.  We compare the
circularization radius to the semi-major axis of the inner binary, for
which we require $a_{\rm c} > a_{\rm in}$, after folding in all
constraints from the requirements for dynamical stability (listed in
the caption of Figure~\ref{fig:fig2}), an outer tertiary that is Roche
lobe-filling, and a dynamically hard outer tertiary orbit.  \simon{Why
  do you consider it important to mention that the orbit is hard, and
  for what environment did you consider this?}  Note that the range of
plotted orbital periods P$_{\rm in}$ corresponding to a contact state
for the inner binary lies outside the range of plotted values for
$P_{\rm in}$ (for components with radii of roughly 1 R$_{\odot}$),
since it does not contribute to constraining the outer orbital
properties).  The thick horizontal solid red line shows the allowed
range of outer semi-major axes, after folding in all of the
aforementioned criteria.  These constraints result in a rather narrow
range of initial conditions for the outer orbit, namely 2.2 $\times$
10$^{2}$ days $\le$ P$_{\rm out}$ $\le$ 1.1 $\times$ 10$^3$ days, for
which the tertiary star overfills its Roche lobe, the triple is (at
least initially) dynamically stable and the accretion stream forms a
gaseous ring around the inner binary.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_M14MSun_outerorbit.pdf}
  \caption{Giant radius as a function of the mass of the core of the
    Roche-lobe filling outer star (dark blue curve).  Here we adopted
    a donor mass of 1.4\,\MSun, but for an 1.2\,\MSun\, donor curve is
    quite similar (see dotted dark-blue curve).  The red squares in
    the curve show the parameters for which we performed more detailed
    gravitational-hydrodynamical simulations (see \S\,\ref{sims}).
    The horizontal dashed line shows the orbital separation of the
    observed twin blue stragger 7782.  The initial triple in which it
    possibly formed must at least have been dynamically stable. The
    minimal orbital separation for the inner binary for which the
    triple is stable is given by the lower green colored curve.
    Donors which are smaller than about 100\,\RSun\, (dark blue curve
    indicated with a$_{\rm stable}$) result in a dynamically unstable
    triple. The minimal core mass associated with a stable triple is
    then indicated by the left-most vertikcal dotted line.  The
    orbital separation at which the donor star overfills its Roche
    lobe is indicated with the light-blue curve. The top curve (brown)
    shows an estimate of the final orbit separation of the outer star,
    and therefore of the final orbit of the white dwarf around the
    inner twin blue straggler.  For core masses $\apgt 0.5$\,\MSun\,
    the final orbital separataion, after mass transfer, is smaller
    than the initial orbit.  Here we adopted an initial inner binary
    mass of (1.0+0.9)\,\MSun\, and a final twin-blue straggler mass of
    (1.4+1.4)\,\MSun.
\label{fig:tertiarymass_vs_size}}
\end{figure}

When adopting a mass of the tertiary star of $m_3 = 1.4$\,\MSun, we
can estimate the initial parameters for the inner binary as well as
the orbit of the outer star after mass transfer. We first calculate
the stellar radius as a function of core mass. In
figure\,\ref{fig:tertiarymass_vs_size} we present this relation
calculated using the {\tt SeBa} stellar evolution code
\cite{1996A&A...309..179P} (dark blue curve).  The interruption in
this curve, aroung a core mass of $m_{\rm core} \sim 0.5$\,\MSun\, is
a result of the evolution along the horizontal branch, where the core
of the star continues to grow but the radius actually shrinks.

Adopting masses for the inner binary $m_1=1.1$\,\MSun\, and
$m_2=0.9$\,\MSun\, we can calculate the outer orbital separation at
the onset of Roche-lobe overflow $a_{\rm out}$, and subsequently the
maximum orbital separation for the inner binary for which the orbit is
stable and a circum binary disk can form. These two limits are
prosented as the light-blue and light green curves in
figure\,\ref{fig:tertiarymass_vs_size}.  The alotted region of
parameter space is then above the dahsed horizontal line and to the
right of the vertical dotted line.

With the adopted parameters, we can also estimate the final orbital
period of the left-over core from the tertiary star.  The change in
orbital separation due to non-conservative mass transfer can be
expressed in terms of the mass of the outer star before and after mass
transfer $m_3$ and $m'_3$, the total mass in the inner binary before
($m_{\rm in}$) and after having accreted ($m'_{\rm in}$) and the
amount of angular momentum lost per unit mass $\eta \simeq 3$
\cite{1995A&A...296..691P}.
\begin{equation}
  {a \over a_0} = \left( {m'_3 m'_{\rm in} \over m_3 m_{\rm in}} \right)^{-2}
  \left( {m'_3 + m'_{\rm in} \over m_3 + m_{\rm in}} \right)^{2\eta + 1}
\end{equation}
In figure\,\ref{fig:tertiarymass_vs_size} we present this relation
with the top brown curve, which subsequently gives an estimate of the
orbital separation of the white dwarf around the twin blue straggler
7782.

@@

\begin{figure}[ht!]
  \includegraphics[width=\columnwidth]{fig_BBSS_gas_M14Msun_A010au_t0002.pdf}
\caption{Top view of the simulated triple system in which the outer
  star over-fills its Roche-lobe (represented by 80000\,sph particles
  and a core particle of 0.4\,\MSun (blue bullet). The two companion
  stars are represented as black bullets (to the right).  The inner
  binary is represented by the yellow and the red bullets for the
  1.1\,\MSun\, primary and 0.9\,\MSun\, secondary stars in a circular
  orbit of 0.1\,au. The 1.4\,\MSun\, giant star is presented to the
  right in a circular orbit of $\sim 1.47$\,au in the plan of the
  inner binary.
\label{fig:topview_at_t0}}
\end{figure}

@@

\begin{figure}[ht!]
  \includegraphics[width=0.49\columnwidth]{fig_BBSS_Lgas_M14Msun_A010au_t3360.pdf}
~  \includegraphics[width=0.51\columnwidth]{fig_BBSS_gas_M14Msun_A010au_t3360.pdf}
\caption{Same as Fig.\,\ref{fig:topview_at_t0} but at an age of $t
  \simeq 1091\,days$ after the start of the simulation. Left panel
  shows the equipotential surface of the triple overplotted with the
  gas distribution, the right panel shows just the gas and the stars
  as bullets.
\label{fig:topview_at_t1000day}}
\end{figure}
